---
title: "Week 12 Lab: Raster Maps of Sea Surface Temperature Anomalies"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: false
execute:
  echo: true
  warning: false
  cache: true
---

# 3.1 Plan the Map Encodings

## Why map anomaly rather than absolute SST?

The principal fill should represent `anom` because the task is to compare departures from each location's usual temperature, not absolute ocean temperature. Absolute SST would be dominated by the normal contrast between warm tropical water and cold high-latitude water, obscuring the interannual signal. The anomaly instead identifies where December conditions were warmer or cooler than the 1971-2000 climatological value for that location and day of year.

## What scale should represent anomaly?

The anomaly requires a diverging scale because it contains meaningful values on both sides of a reference point. Its center is 0 degrees C, meaning the observed SST equals its 1971-2000 climatological value; negative and positive values indicate cooler- and warmer-than-normal conditions respectively. A perceptually neutral color at zero separates these two directions without implying that zero is another low or high value.

## Why must both panels use identical limits?

Both panels must use identical limits so that the same color represents the same anomaly magnitude in each December. Independently rescaled panels could make a weak pattern appear as intense as a strong one and would undermine comparison of both intensity and spatial extent. A common symmetric range calculated from the pooled open-ocean data provides an honest basis for comparison.

# 3.2 Start coding

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(stars)
library(scales)
library(ggnewscale)
library(units)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "100%",
  autodep = TRUE
)

options(dplyr.summarise.inform = FALSE)
```

# 3.3 Import the daily grids as `stars` data cubes

```{r}
#| label: import-data

read_december <- function(path, expected_year) {
  files <- sort(list.files(path, pattern = "[.]nc$", full.names = TRUE))
  dates <- as.Date(
    stringr::str_extract(basename(files), "[0-9]{8}"),
    format = "%Y%m%d"
  )

  stopifnot(
    length(files) == 31L,
    !anyNA(dates),
    isTRUE(all.equal(dates, seq(as.Date(sprintf("%d-12-01", expected_year)),
                         by = "day", length.out = 31L)))
  )

  cube <- read_stars(
    files,
    sub = c("anom", "ice"),
    along = "time",
    proxy = FALSE,
    quiet = TRUE
  )

  stopifnot(identical(names(st_dimensions(cube)), c("x", "y", "zlev", "time")))

  cube <- cube[, , , 1, , drop = TRUE]
  cube <- st_set_dimensions(cube, "time", values = dates, names = "time")
  cube
}

dec_2009 <- read_december(
  "oisst_daily/el_nino_2009-12",
  expected_year = 2009L
)

dec_2010 <- read_december(
  "oisst_daily/la_nina_2010-12",
  expected_year = 2010L
)
```

## 3.3 Import verifier

```{r}
#| label: verify-import

verify_daily_cube <- function(x, year) {
  expected_dates <- seq(
    as.Date(sprintf("%d-12-01", year)),
    by = "day",
    length.out = 31L
  )

  checks <- c(
    attributes = identical(names(x), c("anom", "ice")),
    dimension_names = identical(names(st_dimensions(x)), c("x", "y", "time")),
    dimension_lengths = identical(unname(dim(x)), c(1440L, 720L, 31L)),
    chronological_dates = isTRUE(all.equal(
      as.Date(st_get_dimension_values(x, "time")),
      expected_dates
    ))
  )

  stopifnot(all(checks))

  tibble(
    year = year,
    attributes = paste(names(x), collapse = ", "),
    dimensions = paste(names(dim(x)), dim(x), sep = "=", collapse = ", "),
    first_day = min(as.Date(st_get_dimension_values(x, "time"))),
    last_day = max(as.Date(st_get_dimension_values(x, "time"))),
    all_checks_pass = all(checks)
  )
}

bind_rows(
  verify_daily_cube(dec_2009, 2009L),
  verify_daily_cube(dec_2010, 2010L)
)
```

# 3.4 Establish the coordinate reference system

```{r}
#| label: assign-crs

bbox_before_2009 <- st_bbox(dec_2009)
bbox_before_2010 <- st_bbox(dec_2010)
resolution_before_2009 <- st_res(dec_2009, absolute = FALSE)
resolution_before_2010 <- st_res(dec_2010, absolute = FALSE)

dec_2009 <- st_set_crs(dec_2009, 4326)
dec_2010 <- st_set_crs(dec_2010, 4326)

crs_checks <- tibble(
  year = c(2009L, 2010L),
  correct_crs = c(
    st_crs(dec_2009) == st_crs(4326),
    st_crs(dec_2010) == st_crs(4326)
  ),
  bbox_unchanged = c(
    isTRUE(all.equal(as.numeric(st_bbox(dec_2009)), as.numeric(bbox_before_2009))),
    isTRUE(all.equal(as.numeric(st_bbox(dec_2010)), as.numeric(bbox_before_2010)))
  ),
  resolution_unchanged = c(
    isTRUE(all.equal(st_res(dec_2009, absolute = FALSE), resolution_before_2009)),
    isTRUE(all.equal(st_res(dec_2010, absolute = FALSE), resolution_before_2010))
  )
)

stopifnot(
  all(crs_checks$correct_crs),
  all(crs_checks$bbox_unchanged),
  all(crs_checks$resolution_unchanged),
  isTRUE(all.equal(
    as.numeric(st_bbox(dec_2009)),
    c(0, -90, 360, 90)
  )),
  isTRUE(all.equal(
    as.numeric(st_bbox(dec_2010)),
    c(0, -90, 360, 90)
  ))
)

crs_checks
st_bbox(dec_2009)
```

# 3.5 Calculate monthly cell summaries

```{r}
#| label: monthly-summaries

make_month_summary <- function(december) {
  mean_anom <- st_apply(
    december["anom"],
    c("x", "y"),
    mean,
    na.rm = TRUE
  )
  names(mean_anom) <- "mean_anom"

  mean_ice <- st_apply(
    december["ice"],
    c("x", "y"),
    function(values) mean(replace(values, is.na(values), 0))
  )
  names(mean_ice) <- "mean_ice"

  summary <- c(mean_anom, mean_ice) |>
    mutate(
      land = is.nan(mean_anom) & mean_ice == 0,
      mean_ice = if_else(land, NA_real_, mean_ice)
    )

  summary[c("land", "mean_anom", "mean_ice")]
}

summary_2009 <- make_month_summary(dec_2009)
summary_2010 <- make_month_summary(dec_2010)
```

## 3.5 Monthly-summary verifier

```{r}
#| label: verify-monthly-summaries

verify_month_summary <- function(summary, daily, year) {
  land <- summary[["land"]]
  ice <- summary[["mean_ice"]]
  non_land_ice <- ice[!land]

  checks <- c(
    dimensions = identical(unname(dim(summary)), c(1440L, 720L)),
    no_time = !"time" %in% names(st_dimensions(summary)),
    attributes = identical(names(summary), c("land", "mean_anom", "mean_ice")),
    land_complete = !anyNA(land),
    ice_missing_exactly_on_land = identical(is.na(ice), land),
    ice_in_unit_interval = all(non_land_ice >= 0 & non_land_ice <= 1),
    bbox_unchanged = isTRUE(all.equal(st_bbox(summary), st_bbox(daily))),
    resolution_unchanged = isTRUE(all.equal(st_res(summary), st_res(daily))),
    crs_unchanged = st_crs(summary) == st_crs(daily)
  )

  stopifnot(all(checks))

  tibble(
    year = year,
    dimensions = paste(dim(summary), collapse = " x "),
    attributes = paste(names(summary), collapse = ", "),
    land_cells = sum(land),
    all_checks_pass = all(checks)
  )
}

bind_rows(
  verify_month_summary(summary_2009, dec_2009, 2009L),
  verify_month_summary(summary_2010, dec_2010, 2010L)
)
```

# 3.6 Derive surface classes

```{r}
#| label: surface-classes

surface_levels <- c("land", "sea ice", "open ocean")

add_surface <- function(summary) {
  land <- summary[["land"]]
  sea_ice <- !land & (
    summary[["mean_ice"]] >= 0.15 |
      is.na(summary[["mean_anom"]])
  )

  stopifnot(!anyNA(sea_ice))

  surface_text <- array(
    "open ocean",
    dim = dim(land),
    dimnames = dimnames(land)
  )
  surface_text[sea_ice] <- "sea ice"
  surface_text[land] <- "land"

  surface <- factor(surface_text, levels = surface_levels)
  dim(surface) <- dim(land)
  dimnames(surface) <- dimnames(land)
  summary[["surface"]] <- surface
  summary
}


summary_2009 <- add_surface(summary_2009)
summary_2010 <- add_surface(summary_2010)
```

## 3.6 Surface-class verifier

```{r}
#| label: verify-surface-classes

verify_surface <- function(summary, year) {
  surface <- as.character(summary[["surface"]])
  land <- as.vector(summary[["land"]])
  anomaly <- as.vector(summary[["mean_anom"]])
  ice <- as.vector(summary[["mean_ice"]])

  checks <- c(
    valid_classes = all(surface %in% surface_levels),
    no_missing_class = !anyNA(surface),
    threshold_is_ice = all(surface[!land & ice >= 0.15] == "sea ice"),
    missing_anomaly_fallback = all(surface[!land & is.na(anomaly)] == "sea ice"),
    open_ocean_has_anomaly = !anyNA(anomaly[surface == "open ocean"]),
    open_ocean_below_threshold = all(ice[surface == "open ocean"] < 0.15)
  )

  stopifnot(all(checks))

  tibble(
    year = year,
    land = sum(surface == "land"),
    sea_ice = sum(surface == "sea ice"),
    open_ocean = sum(surface == "open ocean"),
    all_checks_pass = all(checks)
  )
}

bind_rows(
  verify_surface(summary_2009, 2009L),
  verify_surface(summary_2010, 2010L)
)
```

# 3.7 Combine the monthly summaries along a year dimension

```{r}
#| label: combine-years

dec_summaries <- c(
  summary_2009[c("mean_anom", "surface")],
  summary_2010[c("mean_anom", "surface")],
  along = list(year = c(2009L, 2010L))
)

# c.stars() combines factor labels as a plain array, so restore the factor
# explicitly before categorical nearest-neighbor warping.
combined_surface <- factor(
  as.character(dec_summaries[["surface"]]),
  levels = surface_levels
)
dim(combined_surface) <- dim(dec_summaries[["surface"]])
dimnames(combined_surface) <- dimnames(dec_summaries[["surface"]])
dec_summaries[["surface"]] <- combined_surface

combine_checks <- c(
  dimension_names = identical(names(st_dimensions(dec_summaries)), c("x", "y", "year")),
  dimension_lengths = identical(unname(dim(dec_summaries)), c(1440L, 720L, 2L)),
  year_values = identical(
    as.integer(st_get_dimension_values(dec_summaries, "year")),
    c(2009L, 2010L)
  ),
  attributes = identical(names(dec_summaries), c("mean_anom", "surface")),
  bbox = isTRUE(all.equal(st_bbox(dec_summaries), st_bbox(summary_2009))),
  resolution = isTRUE(all.equal(st_res(dec_summaries), st_res(summary_2009))),
  crs = st_crs(dec_summaries) == st_crs(summary_2009)
)

stopifnot(all(combine_checks))

tibble(
  check = names(combine_checks),
  passed = unname(combine_checks)
)
```

```{r}
#| label: release-intermediate-memory
#| cache: false

rm(dec_2009, dec_2010, summary_2009, summary_2010)
invisible(gc())
```

# 3.8 Draft a first faceted raster map

```{r}
#| label: first-draft-map
#| fig-width: 10
#| fig-height: 8

first_draft <- ggplot() +
  geom_stars(
    data = dec_summaries["mean_anom"],
    aes(fill = mean_anom)
  ) +
  facet_wrap(vars(year), ncol = 1)

first_draft
```

The first draft has six important shortcomings:

1. The default continuous fill is a sequential gradient, so a zero anomaly is neither a neutral midpoint nor clearly separated from negative and positive departures.

2. Its endpoints are not deliberately symmetric around zero, allowing one sign of anomaly to receive more visual range than the other.

3. The longitude-latitude view draws equal-degree cells as equally sized rectangles even though their true area decreases toward the poles, so high-latitude patterns look enlarged relative to comparable low-latitude patterns.

4. Land and sea ice share the default missing-value appearance, preventing the reader from distinguishing two substantively different surface types.

5. The unmasked `mean_anom` can display or scale from values at cells that the final design classifies as sea ice, rather than restricting the quantitative encoding to open ocean.

6. The year-only facet labels and default plot furniture do not identify the ENSO phases, anomaly baseline, unit, source, or intended geographic focus.

# 3.9 Design a diverging fill scale

```{r}
#| label: anomaly-mask-and-limit

surface_text <- as.character(dec_summaries[["surface"]])
open_cells <- surface_text == "open ocean"

dec_summaries[["open_ocean_anom"]] <- dec_summaries[["mean_anom"]]
dec_summaries[["open_ocean_anom"]][!open_cells] <- NA_real_

pooled_abs_anomaly <- abs(as.vector(dec_summaries[["open_ocean_anom"]]))
pooled_q99 <- unname(quantile(
  pooled_abs_anomaly,
  probs = 0.99,
  na.rm = TRUE,
  names = FALSE
))
limit <- ceiling(pooled_q99)

stopifnot(is.finite(pooled_q99), is.finite(limit), limit > 0)

tibble(
  pooled_open_ocean_q99_abs_anomaly_c = pooled_q99,
  shared_rounded_limit_c = limit
)
```

```{r}
#| label: diverging-map
#| fig-width: 10
#| fig-height: 8

continuous_anomaly_scale <- function(limit_value, omit_missing = FALSE) {
  scale_fill_distiller(
    name = "Anomaly (°C)",
    type = "div",
    palette = "RdYlBu",
    direction = -1,
    limits = c(-limit_value, limit_value),
    oob = scales::squish,
    na.value = if (omit_missing) NA else "grey50"
  )
}

diverging_map <- ggplot() +
  geom_stars(
    data = dec_summaries["open_ocean_anom"],
    aes(fill = open_ocean_anom)
  ) +
  continuous_anomaly_scale(limit) +
  facet_wrap(vars(year), ncol = 1)

diverging_map
```

# 3.10 Add land and sea ice as surface layers

The anomaly raster must be drawn first and the surface classes above it. Opaque land and sea-ice fills then cover any underlying raster values, while transparent open-ocean cells reveal the anomaly layer. Reversing the order would allow the anomaly raster to paint over land or sea ice and would make the categorical mask visually unreliable.

```{r}
#| label: surface-overlay-map
#| fig-width: 10
#| fig-height: 8

surface_scale <- function() {
  scale_fill_manual(
    name = "Surface",
    values = c(
      "land" = "#8C8C82",
      "sea ice" = "#FFFFFF",
      "open ocean" = "transparent"
    ),
    breaks = c("land", "sea ice"),
    labels = c("Land", "Sea ice (≥15% concentration)"),
    drop = FALSE,
    na.translate = FALSE
  )
}

surface_overlay_map <- ggplot() +
  geom_stars(
    data = dec_summaries["open_ocean_anom"],
    aes(fill = open_ocean_anom)
  ) +
  continuous_anomaly_scale(limit) +
  new_scale_fill() +
  geom_stars(
    data = dec_summaries["surface"],
    aes(fill = surface)
  ) +
  surface_scale() +
  facet_wrap(vars(year), ncol = 1)

surface_overlay_map
```

The surface rules classify every non-land cell with a missing monthly anomaly as sea ice before assigning the remaining cells to open ocean. Consequently, a cell can become open ocean only when its monthly anomaly is available and its mean ice concentration is below 0.15. Because `open_ocean_anom` copies `mean_anom` exactly at those cells, the transparent surface overlay cannot expose a missing anomaly if the rules were implemented correctly.

## 3.10 Anomaly-mask verifier

```{r}
#| label: verify-anomaly-mask

surface_vector <- as.character(dec_summaries[["surface"]])
mean_vector <- as.vector(dec_summaries[["mean_anom"]])
open_vector <- as.vector(dec_summaries[["open_ocean_anom"]])
is_open <- surface_vector == "open ocean"

open_ocean_mismatches <- sum(
  is.na(open_vector[is_open]) |
    open_vector[is_open] != mean_vector[is_open]
)
non_ocean_nonmissing <- sum(!is.na(open_vector[!is_open]))

stopifnot(
  open_ocean_mismatches == 0L,
  non_ocean_nonmissing == 0L
)

tibble(
  open_ocean_value_mismatches = open_ocean_mismatches,
  non_ocean_nonmissing_mask_values = non_ocean_nonmissing
)
```

# 3.11 Compare equal-area world projections

The `lwgeom` package supplies the geodetic segmentization engine required by this installation of `sf`, allowing the boundary edges to be densified with a maximum segment length expressed in degrees.

```{r}
#| label: projection-functions

projection_crs <- c(
  "Equal Earth" = paste(
    "+proj=eqearth +lon_0=180 +ellps=WGS84",
    "+R_A +units=m +no_defs +type=crs"
  ),
  "Mollweide" = paste(
    "+proj=moll +lon_0=180 +ellps=WGS84",
    "+R_A +units=m +no_defs +type=crs"
  )
)

world_longitudes <- function(epsilon = 1e-4, waypoint_degrees = 5) {
  c(
    epsilon,
    seq(waypoint_degrees, 360 - waypoint_degrees, by = waypoint_degrees),
    360 - epsilon
  )
}

make_lonlat_boundary <- function(
    epsilon = 1e-4,
    latitude = 89.5,
    waypoint_degrees = 5,
    max_segment_degrees = 0.25) {
  longitudes <- world_longitudes(epsilon, waypoint_degrees)

  bottom <- cbind(longitudes, -latitude)
  right <- cbind(360 - epsilon, c(-latitude, latitude))
  top <- cbind(rev(longitudes), latitude)
  left <- cbind(epsilon, c(latitude, -latitude))

  ring <- rbind(
    bottom,
    right[-1, , drop = FALSE],
    top[-1, , drop = FALSE],
    left[-1, , drop = FALSE]
  )

  boundary <- st_sfc(
    st_polygon(list(unname(ring))),
    crs = "OGC:CRS84"
  )

  st_segmentize(
    boundary,
    dfMaxLength = set_units(max_segment_degrees, "degree")
  )
}

make_lonlat_equator <- function(
    epsilon = 1e-4,
    waypoint_degrees = 5,
    max_segment_degrees = 0.25) {
  equator <- st_sfc(
    st_linestring(cbind(
      world_longitudes(epsilon, waypoint_degrees),
      0
    )),
    crs = "OGC:CRS84"
  )

  st_segmentize(
    equator,
    dfMaxLength = set_units(max_segment_degrees, "degree")
  )
}

make_lonlat_ocean_labels <- function() {
  st_as_sf(
    tibble(
      label = c("Pacific Ocean", "Atlantic Ocean", "Indian Ocean"),
      longitude = c(200, 330, 80),
      latitude = c(0, 10, -18)
    ),
    coords = c("longitude", "latitude"),
    crs = "OGC:CRS84"
  )
}

restore_surface_factor <- function(x) {
  surface <- factor(
    as.character(x[["surface"]]),
    levels = surface_levels
  )
  dim(surface) <- dim(x[["surface"]])
  dimnames(surface) <- dimnames(x[["surface"]])
  x[["surface"]] <- surface
  x
}

project_summary <- function(source, crs_definition, projection_name) {
  y_values <- as.numeric(
    st_get_dimension_values(source, "y", where = "center")
  )
  keep <- which(abs(y_values) <= 89.5)
  trimmed <- source[, , keep, , drop = FALSE]
  crs <- st_crs(crs_definition)

  target <- st_as_stars(
    st_bbox(st_transform(trimmed, crs)),
    dx = 50000,
    dy = 50000
  )

  warped <- st_warp(trimmed, target, method = "near")
  boundary <- st_transform(make_lonlat_boundary(), crs)
  warped <- st_crop(
    warped,
    boundary,
    crop = FALSE,
    as_points = TRUE
  )
  warped <- restore_surface_factor(warped)

  list(
    name = projection_name,
    crs = crs,
    raster = warped,
    boundary = boundary,
    equator = st_transform(make_lonlat_equator(), crs),
    ocean_labels = st_transform(make_lonlat_ocean_labels(), crs)
  )
}

projected_lines_by_year <- function(x, years = c(2009L, 2010L)) {
  geometry <- if (inherits(x, "sfc")) x else st_geometry(x)
  coordinates <- st_coordinates(geometry)
  part_columns <- intersect(c("L1", "L2", "L3"), colnames(coordinates))

  part <- if (length(part_columns) == 0L) {
    rep("1", nrow(coordinates))
  } else {
    do.call(
      paste,
      c(as.data.frame(coordinates[, part_columns, drop = FALSE]), sep = "_")
    )
  }

  base <- tibble(
    x = coordinates[, 1],
    y = coordinates[, 2],
    part = part
  )

  map_dfr(years, function(year_value) {
    base |>
      mutate(
        year = year_value,
        group = paste(year_value, part, sep = "_")
      )
  })
}

projected_points_by_year <- function(x, years = c(2009L, 2010L)) {
  coordinates <- st_coordinates(x)
  base <- bind_cols(
    st_drop_geometry(x),
    tibble(x = coordinates[, 1], y = coordinates[, 2])
  )

  map_dfr(years, function(year_value) {
    base |> mutate(year = year_value)
  })
}


year_labeller <- as_labeller(c(
  "2009" = "December 2009 (El Niño)",
  "2010" = "December 2010 (La Niña)"
))

build_continuous_map <- function(
    projected,
    title,
    subtitle = NULL,
    caption = NULL,
    label_oceans = FALSE) {
  plot <- ggplot() +
    geom_stars(
      data = projected$raster["open_ocean_anom"],
      aes(fill = open_ocean_anom),
      na.action = na.omit
    ) +
    continuous_anomaly_scale(limit, omit_missing = TRUE) +
    new_scale_fill() +
    geom_stars(
      data = projected$raster["surface"],
      aes(fill = surface),
      na.action = na.omit
    ) +
    surface_scale() +
    geom_path(
      data = projected_lines_by_year(projected$equator),
      aes(x = x, y = y, group = group),
      color = "grey20",
      alpha = 0.25,
      linewidth = 0.2,
      inherit.aes = FALSE,
      show.legend = FALSE
    )

  if (label_oceans) {
    plot <- plot +
      geom_label(
        data = projected_points_by_year(projected$ocean_labels),
        aes(x = x, y = y, label = label),
        fill = alpha("white", 0.55),
        label.size = 0,
        color = "grey20",
        fontface = "italic",
        size = 3,
        inherit.aes = FALSE,
        show.legend = FALSE
      )
  }

  plot +
    geom_path(
      data = projected_lines_by_year(projected$boundary),
      aes(x = x, y = y, group = group),
      color = "grey25",
      linewidth = 0.3,
      inherit.aes = FALSE,
      show.legend = FALSE
    ) +
    facet_wrap(
      vars(year),
      ncol = 1,
      labeller = labeller(year = year_labeller)
    ) +
    coord_equal(expand = FALSE, clip = "on") +
    labs(
      title = title,
      subtitle = subtitle,
      caption = caption
    ) +
    theme_void(base_size = 10) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey96", color = NA),
      panel.spacing = grid::unit(4, "pt"),
      legend.position = "right",
      legend.key = element_rect(fill = "grey97", color = "grey80")
    )
}
```

```{r}
#| label: project-both-grids

equal_earth <- project_summary(
  dec_summaries,
  projection_crs[["Equal Earth"]],
  "Equal Earth"
)

mollweide <- project_summary(
  dec_summaries,
  projection_crs[["Mollweide"]],
  "Mollweide"
)

projection_checks <- tibble(
  projection = c(equal_earth$name, mollweide$name),
  crs_is_projected = c(
    !st_is_longlat(equal_earth$raster),
    !st_is_longlat(mollweide$raster)
  ),
  year_length = c(
    dim(equal_earth$raster)[["year"]],
    dim(mollweide$raster)[["year"]]
  ),
  surface_is_factor = c(
    is.factor(equal_earth$raster[["surface"]]),
    is.factor(mollweide$raster[["surface"]])
  )
)

stopifnot(
  all(projection_checks$crs_is_projected),
  all(projection_checks$year_length == 2L),
  all(projection_checks$surface_is_factor)
)

projection_checks
```

## Equal Earth

```{r}
#| label: equal-earth-comparison
#| fig-width: 10
#| fig-height: 9

equal_earth_comparison <- build_continuous_map(
  equal_earth,
  title = "Equal Earth Projection Comparison",
  subtitle = "Open-ocean anomaly with land and sea-ice overlay"
)

equal_earth_comparison
```

## Mollweide

```{r}
#| label: mollweide-comparison
#| fig-width: 10
#| fig-height: 9

mollweide_comparison <- build_continuous_map(
  mollweide,
  title = "Mollweide Projection Comparison",
  subtitle = "Open-ocean anomaly with land and sea-ice overlay"
)

mollweide_comparison
```

### Polar representation

Equal Earth retains a finite-width, flattened polar edge, so high-latitude regions do not converge to a single tip. Mollweide tapers the outline to a point at each pole, compressing polar shapes while still preserving their areas.

### Equatorial Pacific clarity

Both projections use a 180-degree central meridian, so neither splits the equatorial Pacific at a panel edge. Mollweide makes the continuous equatorial cool and warm bands especially easy to follow because its widest horizontal span occurs exactly at the equator.

### Use of rectangular panel space

Equal Earth uses more of the rectangular panel because its high-latitude outline retains width, whereas Mollweide's ellipse leaves larger blank corners. The rendered Equal Earth panels therefore devote slightly more ink to polar and mid-latitude regions.

### Prominence of the contrast between Decembers

With identical data and limits, the December contrast is more visually prominent in Mollweide along the central and eastern equatorial Pacific. Its geometry gives the main ENSO band a strong, uninterrupted central axis, while the deliberately spare frame keeps attention on the sign reversal.

### Projection choice

Both candidates are equal-area, so either supports comparisons of spatial extent. Mollweide is chosen because its broad, 180-degree-centered equatorial span keeps the Pacific signal central and clearest in the rendered comparison. Its elliptical outline also reads immediately as a complete world map. The tradeoff is unused corner space and compression of the poles to points, which is acceptable because the principal analytic focus is equatorial SST anomaly rather than polar shape.

# 3.12 Build the final continuous two-panel map

```{r}
#| label: final-continuous-map
#| fig-width: 10
#| fig-height: 9

final_title <- "Sea Surface Temperature Anomalies During El Niño and La Niña"
final_subtitle <- "Monthly mean departure from the 1971-2000 climatology"
final_caption <- paste(
  "Source: NOAA 0.25° Daily OISST v2.1",
  "(doi:10.25921/RE9P-PT57)."
)

final_continuous <- build_continuous_map(
  mollweide,
  title = final_title,
  subtitle = final_subtitle,
  caption = final_caption,
  label_oceans = TRUE
)

final_continuous
```

# 3.13 Bin the anomaly scale

```{r}
#| label: binned-map-functions

make_anomaly_classes <- function(values, limit_value) {
  edges <- seq(-limit_value, limit_value, by = 1)

  signed_label <- function(value) {
    if_else(value > 0, paste0("+", value), as.character(value))
  }

  bounded_labels <- paste(
    signed_label(edges[-length(edges)]),
    "to",
    signed_label(edges[-1])
  )
  lower_label <- paste("<", signed_label(-limit_value))
  upper_label <- paste(">", signed_label(limit_value))
  ordered_labels <- c(lower_label, bounded_labels, upper_label)

  numeric_values <- as.numeric(values)
  class_text <- rep(NA_character_, length(numeric_values))

  class_text[numeric_values < -limit_value] <- lower_label
  class_text[numeric_values > limit_value] <- upper_label

  inside <- which(
    is.finite(numeric_values) &
      numeric_values >= -limit_value &
      numeric_values <= limit_value
  )

  interval_index <- findInterval(
    numeric_values[inside],
    edges,
    rightmost.closed = TRUE,
    all.inside = TRUE
  )
  interval_index <- pmin(interval_index, length(bounded_labels))
  class_text[inside] <- bounded_labels[interval_index]

  classes <- factor(
    class_text,
    levels = ordered_labels,
    ordered = TRUE
  )
  dim(classes) <- dim(values)
  dimnames(classes) <- dimnames(values)

  base_palette <- scales::brewer_pal(
    type = "div",
    palette = "RdYlBu",
    direction = -1
  )(11)
  colors <- grDevices::colorRampPalette(
    base_palette,
    space = "Lab"
  )(length(ordered_labels))

  list(
    values = classes,
    levels = ordered_labels,
    colors = setNames(colors, ordered_labels)
  )
}

build_binned_map <- function(
    projected,
    title,
    subtitle,
    caption,
    label_oceans = TRUE) {
  classified <- make_anomaly_classes(
    projected$raster[["open_ocean_anom"]],
    limit
  )
  raster <- projected$raster
  raster[["anomaly_class"]] <- classified$values

  plot <- ggplot() +
    geom_stars(
      data = raster["anomaly_class"],
      aes(fill = anomaly_class),
      na.action = na.omit
    ) +
    scale_fill_manual(
      name = "Anomaly (°C)",
      values = classified$colors,
      breaks = classified$levels,
      limits = classified$levels,
      drop = FALSE,
      na.translate = FALSE,
      guide = guide_legend(order = 1)
    ) +
    new_scale_fill() +
    geom_stars(
      data = raster["surface"],
      aes(fill = surface),
      na.action = na.omit
    ) +
    surface_scale() +
    geom_path(
      data = projected_lines_by_year(projected$equator),
      aes(x = x, y = y, group = group),
      color = "grey20",
      alpha = 0.25,
      linewidth = 0.2,
      inherit.aes = FALSE,
      show.legend = FALSE
    )

  if (label_oceans) {
    plot <- plot +
      geom_label(
        data = projected_points_by_year(projected$ocean_labels),
        aes(x = x, y = y, label = label),
        fill = alpha("white", 0.55),
        label.size = 0,
        color = "grey20",
        fontface = "italic",
        size = 3,
        inherit.aes = FALSE,
        show.legend = FALSE
      )
  }

  plot +
    geom_path(
      data = projected_lines_by_year(projected$boundary),
      aes(x = x, y = y, group = group),
      color = "grey25",
      linewidth = 0.3,
      inherit.aes = FALSE,
      show.legend = FALSE
    ) +
    facet_wrap(
      vars(year),
      ncol = 1,
      labeller = labeller(year = year_labeller)
    ) +
    coord_equal(expand = FALSE, clip = "on") +
    labs(title = title, subtitle = subtitle, caption = caption) +
    theme_void(base_size = 10) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      strip.text = element_text(face = "bold", size = 10),
      strip.background = element_rect(fill = "grey96", color = NA),
      panel.spacing = grid::unit(4, "pt"),
      legend.position = "right",
      legend.key = element_rect(fill = "grey97", color = "grey80")
    )
}
```

```{r}
#| label: final-binned-map
#| fig-width: 10
#| fig-height: 9

final_binned <- build_binned_map(
  mollweide,
  title = final_title,
  subtitle = final_subtitle,
  caption = final_caption,
  label_oceans = TRUE
)

final_binned
```

# 3.14 Reflect on the visualization

## Difference between the two Decembers

Along the central and eastern equatorial Pacific, approximately 160 degrees E to 80 degrees W, December 2009 shows a broad positive anomaly band, whereas December 2010 shows an extensive negative band. The 2009 warm signal is continuous across much of the central Pacific and reaches eastward toward the South American coast; the 2010 cool tongue is stronger and more spatially coherent along the same equatorial corridor. Secondary patches in the North Pacific and parts of the Southern Ocean also change sign or intensity, but they are less continuous than the tropical Pacific signal. Because both panels use one scale and one equal-area projection, these statements compare departures from local climatology rather than absolute SST or a causal temperature trend.

## Gains and losses from the open-ocean and sea-ice choices

Restricting the quantitative layer to open ocean keeps the anomaly scale tied to visible ocean-temperature data and prevents land, ice, or missing cells from masquerading as SST information. A single sea-ice class makes the frozen surface easy to locate without letting it compete with the ENSO signal. The tradeoff is that the reader cannot see anomaly values beneath the ice overlay or distinguish a cell just above 15% mean concentration from one with almost complete ice cover. The class also includes non-land missing-anomaly boundary cells, so it cannot support claims about detailed ice concentration, thickness, or change.

## Continuous versus binned anomaly color

For comparing spatial extent, the binned scale is stronger because common threshold colors make the footprints of comparable warm and cool classes easier to trace across panels. It sacrifices within-bin variation and creates an artificial visual boundary between values that may be numerically very close. The continuous scale preserves subtle gradients and local structure, but the boundary of any specified anomaly range is less distinct. The binned version is therefore preferred for extent comparison, while the continuous version remains more informative when fine variation matters.

# 4.1 Reflect on AI Use
Our team used ChatGPT as our web-based AI tool.

One useful suggestion came in Section 3.11 where we had some issues constructing the map plot. We described the problem and our own workaround to ChatGPT, and it introduced st_segmentize() as the right tool to densify a line or polygon up to a maximum segment length instead of us hand-tuning a uniform spacing.

That same response was incomplete as it never mentioned that segmentizing on a geographic CRS also depends on the lwgeom package, which isn't one of the packages in Section 2.2. We only found this out after running the suggested code block so we added a short note before the projection-functions chunk explaining what lwgeom provides and why it's required.

# 4.2 Member’s Role and Contributions
# Task 5.2
- Joshua Dui: Coder
- Benjamin: Navigator
- Mei Yuen: Verifier
- Dannon: Constructive Critic and Reporter
- Danish: AI Liaison